---
title: 'Project 2: Modeling, Testing, and Predicting'
author: "SDS348 Fall 2019"
date: ''
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})

knitr::opts_chunk$set(echo = TRUE, eval = TRUE,fig.align="center",warning=FALSE,message=FALSE,fig.width=8, fig.height=5, linewidth=60)
options(tibble.width = 100,width = 100)
library(tidyverse)



```

```{r eval = F}
library(knitr)
opts_chunk$set(fig.align="center", fig.height=5, message=FALSE, warning=FALSE, fig.width=8, tidy.opts=list(width.cutoff=60),tidy=TRUE)

class_diag<-function(probs,truth){
  
  tab<-table(factor(probs>.5,levels=c("FALSE","TRUE")),truth)
  acc=sum(diag(tab))/sum(tab)
  sens=tab[2,2]/colSums(tab)[2]
  spec=tab[1,1]/colSums(tab)[1]
  ppv=tab[2,2]/rowSums(tab)[2]

  if(is.numeric(truth)==FALSE & is.logical(truth)==FALSE) truth<-as.numeric(truth)-1
  
  #CALCULATE EXACT AUC
  ord<-order(probs, decreasing=TRUE)
  probs <- probs[ord]; truth <- truth[ord]
  
  TPR=cumsum(truth)/max(1,sum(truth)) 
  FPR=cumsum(!truth)/max(1,sum(!truth))
  
  dup<-c(probs[-1]>=probs[-length(probs)], FALSE)
  TPR<-c(0,TPR[!dup],1); FPR<-c(0,FPR[!dup],1)
  
  n <- length(TPR)
  auc<- sum( ((TPR[-1]+TPR[-n])/2) * (FPR[-1]-FPR[-n]) )

  data.frame(acc,sens,spec,ppv,auc)
}
```
# Modeling

## Instructions

A knitted R Markdown document (as a PDF) and the raw R Markdown file (as .Rmd) should both be submitted to Canvas by 11:59pm on 11/24/2019. These two documents will be graded jointly, so they must be consistent (i.e., don’t change the R Markdown file without also updating the knitted document). Knit an html copy too, for later! In the .Rmd file for Project 2, you can copy the first code-chunk into your project .Rmd file to get better formatting. Notice that you can adjust the opts_chunk$set(...) above to set certain parameters if necessary to make the knitting cleaner (you can globally set the size of all plots, etc). I have gone ahead and set a few for you (such as disabling warnings and package-loading messges when knitting)! 

Like before, I envision your written text forming something of a narrative structure around your code/output. All results presented must have corresponding code. Any answers/results/plots etc. given without the corresponding R code that generated the result will not be graded. Furthermore, all code contained in your final project document should work properly. Please do not include any extraneous code or code which produces error messages. (Code which produces warnings is acceptable, as long as you understand what the warnings mean).

## Find data:

Find one dataset with at least 5 variables that wish to use to build model. At least one should be categorical (with 2-5 groups) and at least two should be numeric. Ideally, one of your variables will be binary (if not, you will need to create one by discretizing a numeric, which is workable but less than ideal). You will need a minimum of 40 observations (*at least* 10 observations for every explanatory variable you have, ideally 20+ observations/variable).

It is perfectly fine to use either dataset (or the merged dataset, or a subset of your variables) from Project 1. However, you could also diversify your portfolio a bit by choosing a different dataset to work with (particularly if the variables did not reveal interesting associations in Project 1). The only requirement/restriction is that you may not use data from any examples we have done in class or lab. It would be a good idea to pick more cohesive data this time around (i.e., variables that you actually thing might have a relationship you would want to test). Think more along the lines of your Biostats project.

Again, you can use data from anywhere you want (see bottom for resources)! If you want a quick way to see whether a built-in (R) dataset has binary and/or character (i.e., categorical) variables, check out this list: https://vincentarelbundock.github.io/Rdatasets/datasets.html.
```{R eval=F}
data(package = .packages(all.available = TRUE))
library(boot)
help(melanoma)
melanoma <- data.frame(melanoma)
```


## Guidelines and Rubrc

- **0. (5 pts)** Introduce your dataset and each of your variables (or just your main variables if you have lots) in a paragraph.
```{R eval=F}
#The data set melanoma contains measurements of patients who have been afflicted with melanoma. All 205 patients listed in this dataset have had their tumors removed from the period of 1962 to 1977. The variables included are time(survival time in days since the operations), status(1 means they died from melanoma, 2 means that were still alive, and 3 means that had passed away from unrelated reasons), sex(male or female), age(their age in years when they had the melanoma removed), year(the actual year of the operation), thickness(tumor thickness measured in mm) and ulcer(detection of an ulcer after tumor was removed. )
```

- **1. (15 pts)** Perform a MANOVA testing whether any of your numeric variables (or a subset of them, if including them all doesn't make sense) show a mean difference across levels of one of your categorical variables (3). If they do, perform univariate ANOVAs to find response(s) showing a mean difference across groups (3), and perform post-hoc t tests to find which groups differ (3). Discuss the number of tests you have performed, calculate the probability of at least one type I error (if unadjusted), and adjust the significance level accordingly (bonferroni correction) before discussing significant differences (3). Briefly discuss assumptions and whether or not they are likely to have been met (2).

Answer:It appears that there is a significant mean difference across the groups of the categorical status, based on one of my numeric variables. Alive & death by melanoma, and alive & unrelated deaths differs significantly across status based on time. Alive & death, alive & unrelated death, and melanoma death & unrelated death differs significantly across status based on age. Alive & death by melanoma differs significantly across status based on year. Alive & death by melanoma differs significantly across status based on tumor thickness. Alive & death by melanoma differs significantly across status based on time. The probability that at least a Type 1 error was made is 0.7226104. #The bonferroni adjusted significant level is 0.02890442. After the bonferroni correction, there is now no longer a significant difference between those who died from melanoma & those who died from unrelated causes, based on AGE. The same can be said based on TUMOR THICKNESS and whether or not there was an ULCER. The shape of the lines and the distance between the lines indicates that the data has multivariate normality. Since the values between the groups are similar in value, this shows that the assumption for covariance was met. 
```{R eval=F}
#MANOVA writeup found on WS 15 slide 38
library(dplyr)
library(tidyverse)
library(tidyr)
library(tidyselect)
library(tidyselect)
library(MASS)
library(lmtest)
library(boot)
library(glmnet)
melanoma <- melanoma %>% mutate(status_cat = case_when(status == 1 ~ "melanoma_death", status == 2 ~ "alive", status == 3 ~ "unrelated_death"))
#MANOVA testing
meldiff <- manova(cbind(time, sex, age, year, thickness, ulcer) ~ status_cat, data = melanoma)
summary(meldiff)
summary(aov(time~status_cat, data = melanoma)) #sig
summary(aov(sex~status_cat, data = melanoma))
summary(aov(age~status_cat, data = melanoma)) #sig
summary(aov(year~status_cat, data = melanoma)) #sig
summary(aov(thickness~status_cat, data = melanoma)) #sig
summary(aov(ulcer~status_cat, data = melanoma)) #sig

pairwise.t.test(melanoma$time, melanoma$status_cat, p.adj = "none")
pairwise.t.test(melanoma$age, melanoma$status_cat, p.adj = "none")
pairwise.t.test(melanoma$year, melanoma$status_cat, p.adj = "none")
pairwise.t.test(melanoma$thickness, melanoma$status_cat, p.adj = "none")
pairwise.t.test(melanoma$ulcer, melanoma$status_cat, p.adj = "none")
1 - (1 - 0.05)^25
1 - .95^25 
0.7226104/25

ggplot(melanoma, aes(x = thickness, y = age)) +
 geom_point(alpha = .5) + geom_density_2d(h=500) + coord_fixed() + facet_wrap(~status_cat)
covmats<-melanoma%>%group_by(status_cat)%>%do(covs=cov(.[2:6]))
for(i in 1:3){print(covmats$covs[i])}

```

- **2. (10 pts)** Perform some kind of randomization test on your data (that makes sense). This can be anything you want! State null and alternative hypotheses, perform the test, and interpret the results (7). Create a plot visualizing the null distribution and the test statistic (3).

Answer: I am trying to see if tumor thickness has some sort of relationship with whether there was an ulcer or not, when the tumor was removed
Ho = mean tumor thickness when removed is the same for whether there was an ulcer vs no ulcer
Ha = mean tumor thickness when removed is different for whether there was an ulcer vs no ulcer
Null hypothesis is rejected, there is a difference between thickness and whether an ulcer was there or not. In addition, the p-value of 2.042e-09 if far less than that of 0.05 so this is another indicator that the null hypothesis was rejected.
```{R eval=F}
melanoma <- melanoma %>% mutate(ulcer_cond = case_when(ulcer == 1 ~ "ulcer_yes", ulcer == 0 ~ "ulcer_no"))

melanoma %>% group_by(ulcer_cond) %>% summarize(mean_thick = mean(thickness)) %>% summarize(diff(mean_thick))
	
rand_dist<-vector()
for(i in 1:5000){
new_melanoma<-data.frame(thickness=sample(melanoma$thickness),ulcer_cond=melanoma$ulcer_cond)
rand_dist[i]<-mean(new_melanoma[new_melanoma$ulcer_cond=="ulcer_yes",]$thickness)-
 mean(new_melanoma[new_melanoma$ulcer_cond=="ulcer_no",]$thickness)}

mean(rand_dist>2.525029)*2

{hist(rand_dist,main="", ylab=""): abline(v = 2.525029, col = "red")}

t.test(melanoma$thickness~melanoma$ulcer_cond)
 
```

- **3. (35 pts)** Build a linear regression model predicting one of your response variables from at least 2 other variables, including their interaction. Mean-center any numeric variables involved in the interaction.

    - Interpret the coefficient estimates (do not discuss significance) (10)
    - Plot the regression using `ggplot()`. If your interaction is numeric by numeric, refer to code near the end of WS15 to make the plot. If you have 3 or more predictors, just chose two to plot for convenience. (7)
    - Check assumptions of linearity, normality, and homoskedasticity either graphically or using a hypothesis test (3)
    - Regardless, recompute regression results with robust standard errors via `coeftest(..., vcov=vcovHC(...))`. Discuss significance of results, including any changes from before/after robust SEs if applicable. (7)
    - What proportion of the variation in the outcome does your model explain? (3)
Answer: 
The intercept value is 2171.872 which indicates the survival of the patient in days when age_c and thickness_c are controlled for and equal to 0. The estimate value is -17.506. Controlling for thickness_c by setting it equal to 0, there is a 17.506 decrease in days of survival after the operation for every increase in age_c in years. Controlling for age_c, there is a 55.816 decrease in days of survival after the operation for every 1mm increase in thickness_c. The interaction estimate is -1.828. This indicates that for every 1 year increase in age_c, the slope for thickness_c vs survival time in days decreases by 1.828. The assumption for linearity and homoskedacicity is met because the distribution between the top and bottom of the 0 mark of residuals is evenly spread. Normality - the graph appears to be normally distributed. It appears that robust standard errors for mean centered age, mean centered thickness and mean centered values for the interaction of thickness and age, were all larger than the unrobust standard errors. 0.1277 is the proportion of variation of time(days) explained by the overall model which includes all the predictors at the same time
    ```{R eval=F}
#predict survival time from thickness and age
library(sandwich)
install.packages("interactions")  
library(interactions)
    melanoma$age_c <- melanoma$age - mean(melanoma$age, na.rm = T)
melanoma$thickness_c <- melanoma$thickness - mean(melanoma$thickness, na.rm = T)
melanoma$time_c <- melanoma$time - mean(melanoma$time, na.rm = T)
melanoma$year_c <- melanoma$year - mean(melanoma$year, na.rm = T)
fit <- lm(time ~ age_c*thickness_c, data = melanoma,)
interact_plot(fit, thickness_c, age_c, data=melanoma)
summary(fit)
resids<-fit$residuals
fitvals<-fit$fitted.values
ggplot()+geom_point(aes(fitvals,resids))+geom_hline(yintercept=0, color='red')
ggplot()+geom_histogram(aes(resids), bins=20)
#computing with robust standard errors
bptest(fit)
coeftest(fit)[,1:2] 
coeftest(fit, vcov=vcovHC(fit))[,1:2] 
summary(fit)


```

- **4. (5 pts)** Rerun same regression model (with interaction), but this time compute bootstrapped standard errors. Discuss any changes you observe in SEs and p-values using these SEs compared to the original SEs and the robust SEs)
Answer: 
It appears that the bootstrapped standard errors are higher than the original/unrobust standard errors but greater than the robust standard errors. The bootstrap SE for age_c, thickness_c and interaction between age_c & thickness_c is higher than that of the original SE but lower than that of the robust SE. This means that the p-values are also in between that of the original and robust SE. 

    ```{R eval=F}
samp_distn<-replicate(5000, {
 boot_dat<-boot_dat<-melanoma[sample(nrow(melanoma),replace=TRUE),]
 fit2<-lm(time ~ age_c * thickness_c, data=boot_dat)
 coef(fit2)
})
## Estimated SEs
samp_distn%>%t%>%as.data.frame%>%summarize_all(sd)
#Empirical 95% CI
#samp_distn%>%t%>%as.data.frame%>%gather%>%group_by(key)%>%
#summarize(lower=quantile(value,.025), upper=quantile(value,.975))
coeftest(fit)[,1:2] 
coeftest(fit, vcov=vcovHC(fit))[,1:2]
```

- **5. (40 pts)** Perform a logistic regression predicting a binary categorical variable (if you don't have one, make/get one) from at least two explanatory variables (interaction not necessary). 

    - Interpret coefficient estimates in context (10)
    - Report a confusion matrix for your logistic regression (2)
    - Compute and discuss the Accuracy, Sensitivity (TPR), Specificity (TNR), and Recall (PPV) of your model (5)
    - Using ggplot, plot density of log-odds (logit) by your binary outcome variable (3)
    - Generate an ROC curve (plot) and calculate AUC (either manually or with a package); interpret (10)
    - Perform 10-fold (or repeated random sub-sampling) CV and report average out-of-sample Accuracy, Sensitivity, and Recall (10)
  
Answer: For every one mm increase in tumor thickness, the odds of having an ulcer are multiplied by 1.582095. The odds of having an ulcer increases by 58.21% for every additional mm increase in thickness. For every one year increase that passes, the odds of having an ulcer are multiplied by 1.025818. In other words, the odds of having an ulcer increases by 2.55% for every additional year that passes. Accuracy = 0.7463415 is the proportion of cases that were correctly classified. Sensitivity = 0.5666667 is the proportion of cases, where an ulcer was present, that were correctly predicted out of all the cases in which the ulcer was present. Specificity = 0.8869565 is the proportion of cases, where an ulcer was not present, that were correctly predicted out of all the cases in which the ulcer was not present. PPV = 0.796875 is the proportion of cases, where an ulcer was present, where the prediction was correctly identified as a patient having an ulcer, out of all cases where the test result indicated that there was an ulcer(test result positive) 

    ```{R eval=F}
#predicting whether someone will have an ulcer or not based on tumor thickness and gender
library(lmtest)
library(boot)
library(glmnet)
library(plotROC)

fit3 <- glm(ulcer ~ thickness + year, data = melanoma, family = binomial(link="logit"))
summary(fit3)
exp(0.45875)
#1.582095
exp(0.02549)
#1.02581
#confusion matrix
prob <- predict(fit3, type = "response") #get predicted probabilites for full data
pred <- ifelse(prob > 0.5, 1, 0) #get predictions (if you want)
rbind(prob, pred, truth = melanoma$ulcer)[, 1:10] %>% round(3)
table(predict= as.numeric(prob>.5), truth = melanoma$ulcer)%>%addmargins
#classification diagnostics
class_diag(prob, melanoma$ulcer)

#logit plot
odds<-function(p)p/(1-p)
p<-seq(0,1,by=.1)
cbind(p, odds=odds(p))%>%round(4)
logit<-function(p)log(odds(p))
cbind(melanoma$ulcer, odds=odds(p),logit=logit(p))%>%round(4)
melanoma$logit<- predict(fit3)
ggplot(melanoma) + geom_density(aes(logit, fill=as.factor(ulcer)))

#This is the ROC cure and the calculations for area under the curve 
install.packages("pROC")
library(pROC)

ggplot(melanoma,aes(ulcer,pred,color=thickness))+geom_line()
ROCplot<-ggplot(melanoma)+geom_roc(aes(d=sex,m=pred), n.cuts=0)+
 geom_segment(aes(x=0,xend=1,y=0,yend=1),lty=2)
ROCplot
calc_auc(ROCplot)
#0.6373317	

#This the 10 fold cv subsetting
set.seed(1234)
fraction<-0.5 #choose proportion of rows to train
train_n<-floor(fraction*nrow(melanoma)) #number of rows to train
iter<-500 #number of iterations
diags<-NULL
for(i in 1:iter){
 ## Create training and test sets
 train_index<-sample(1:nrow(melanoma),size=train_n)
 train<-melanoma[train_index,]
 test<-melanoma[-train_index,]
 truth<-test$ulcer
 ## Train model on training set
 fit4<-glm(ulcer ~ thickness + year,data=train,family="binomial")
 probs<-predict(fit4,newdata = test,type="response")
 ## Test model on test set (save results)
 diags<-rbind(diags,class_diag(probs,truth))
}
apply(diags,2,mean) #average across all k results

```

- **6. (10 pts)** Choose one variable you want to predict (can be one you used from before; either binary or continuous) and run a LASSO regression inputting all the rest of your variables as predictors. Choose lambda to give the simplest model whose accuracy is near that of the best (i.e., `lambda.1se`). Discuss which variables are retained. Perform 10-fold CV using this model: if response in binary, compare model's out-of-sample accuracy to that of your logistic regression in part 5; if response is numeric, compare the residual standard error (at the bottom of the summary output, aka RMSE): lower is better fit!

Answer: It appears that the variables thickness, status, and time were retained and sex was taken out. The Auc for number 5 was 0.6373317, compared to that of the increased auc value of 0.7919184.

```{R eval=F}
fit5<-glm(ulcer ~ thickness + year  + age + sex + status + time, data=melanoma, family = "binomial")
a <- model.matrix(fit5)[, -1]
b <- as.matrix(melanoma$ulcer)
cv <- cv.glmnet(a, b)
lasso <- glmnet(a, b, lambda = cv$lambda.1se,)
coef(lasso)
set.seed(1234)
k = 10
mel1 <- melanoma[sample(nrow(melanoma)), ]
folds <- cut(seq(1:nrow(melanoma)), breaks = k, labels = F)
diags <- NULL
for (i in 1:k) {
train <- mel1[folds != i, ]
test <- mel1[folds == i, ]
truth <- test$ulcer
fit6<-glm(ulcer ~ thickness + status + time, data=train, family = "binomial")
probs2 <- predict(fit6, newdata = test, type = "response")
diags <- rbind(diags, class_diag(probs2, truth))
}
diags%>%summarise_all(mean)

 

```

## Where do I find data again?

You can choose ANY datasets you want that meet the above criteria for variables and observations. You can make it as serious as you want, or not, but keep in mind that you will be incorporating this project into a portfolio webpage for your final in this course, so choose something that really reflects who you are, or something that you feel will advance you in the direction you hope to move career-wise, or something that you think is really neat, or whatever. On the flip side, regardless of what you pick, you will be performing all the same tasks, so it doesn't end up being that big of a deal.

If you are totally clueless and have no direction at all, log into the server and type 

```{R eval=F}
data(package = .packages(all.available = TRUE))
```

This will print out a list of **ALL datasets in ALL packages** installed on the server (a ton)! Scroll until your eyes bleed! Actually, do not scroll that much... To start with something more manageable, just run the command on your own computer, or just run `data()` to bring up the datasets in your current environment. To read more about a dataset, do `?packagename::datasetname`. 

If it is easier for you, and in case you don't have many packages installed, a list of R datasets from a few common packages (also downloadable in CSV format) is given at the following website: https://vincentarelbundock.github.io/Rdatasets/datasets.html.

- A good package to download for fun/relevant data is `fivethiryeight`. 
Run `install.packages("fivethirtyeight"),` load the packages with `library(fivethirtyeight)`, run `data()`, and then scroll down to view the datasets. Here is an online list of all 127 datasets (with links to the 538 articles). Lots of sports, politics, current events, etc.

- If you have already started to specialize (e.g., ecology, epidemiology) you might look at discipline-specific R packages (vegan, epi, respectively). We will be using some tools from these packages later in the course, but they come with lots of data too, which you can explore according to the directions above

- However, you *emphatically DO NOT* have to use datasets available via R packages! In fact, I would much prefer it if you found the data from completely separate sources and brought them together (a much more realistic experience in the real world)! You can even reuse data from your SDS328M project, provided it shares a variable in common with other data which allows you to  merge the two together (e.g., if you still had the timestamp, you could look up the weather that day: https://www.wunderground.com/history/). If you work in a research lab or have access to old data, you could potentially merge it with new data from your lab!

- Here is a curated list of interesting datasets (read-only spreadsheet format): https://docs.google.com/spreadsheets/d/1wZhPLMCHKJvwOkP4juclhjFgqIY8fQFMemwKL2c64vk/edit

- Here is another great compilation of datasets: https://github.com/rfordatascience/tidytuesday

- Here is the UCI Machine Learning Repository: https://archive.ics.uci.edu/ml/index.php

    - See also https://en.wikipedia.org/wiki/List_of_datasets_for_machine-learning_research#Biological_data

- Here is another good general place to look: https://www.kaggle.com/datasets

- To help narrow your search down or to see interesting variable ideas, check out https://www.tylervigen.com/spurious-correlations. This is the spurious correlations website, and it is fun, but if you look at the bottom of each plot you will see sources for the data. This is a good place to find very general data (or at least get a sense of where you can scrape data together from)!

- If you are interested in medical data, check out www.countyhealthrankings.org

- If you are interested in scraping UT data, they make *loads* of data public (e.g., beyond just professor CVs and syllabi). Check out all the data that is available in the statistical handbooks: https://reports.utexas.edu/statistical-handbook

##### Broader data sources:

[Data.gov](www.data.gov) 186,000+ datasets!

[Social Explorer](Social Explorer) is a nice interface to Census and American Community Survey data (more user-friendly than the government sites). May need to sign up for a free trial.

[U.S. Bureau of Labor Statistics](www.bls.gov)

[U.S. Census Bureau](www.census.gov)

[Gapminder](www.gapminder.org/data), data about the world.

...





